Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-07 paper/main.tex
typo
some typos

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
% The purpose of this section is to build some intuition that should
% make the following sections easy to read.

@@

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
quantity we are interested in.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that such that $(S, \state) \Downarrow_n \state'$.

To obtain a compositional analysis we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre and post conditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  Thus,
we use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatably increments a binary counter that is
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks that are
executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.  A
naive analysis of the algorithm would lead to a quadratic bound since
there can be a linear number of updates in the inner loop.

@@

See \pref{sec:logic} for more a formal definition of the logic and
additional example derivation.

@@

To enable automation, we fix the shape of the potential functions
$\Phi$ so that it becomes possible to use \emph{linear programming}
(LP) to compute a derivation in the logic.  If we assume for now that
a program state $\state$ simply maps variable to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} \iffull{and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$. Note
that there are no restrictions on the signs of the input variables in
the examples.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variables for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear inequalities on
these variables that enforce that the potential annotations are sound.
For instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

@@

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource cost is closely related to the
number of times the test \lstinline{a[i] > m} is true during the
execution.
%
If we define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
the number of ``maximum candidates'' in the array \lstinline{a} seen
by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
So any automated tool would at best derive the linear bound $2 \cdot
N$ for that program.  But with the expressivity of our logic it is
possible to use the previous set cardinal directly and precisely tie
the bound to the initial contents of the array.
%
The non-trivial part of this derivation is finding the loop invariant
$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
When the condition \lstinline{a[i] > m} is true, we know that we
encountered a ``maximum so far'' because $m$ is a maximal element of
\lstinline{a[$0 \dots i$]}, thus $A(i) = 1 + A(i+1)$ and we get one
potential unit to pay for the assignment.  In the other case, no
maximum so far is encountered so $A(i) = A(i+1)$.
% As a side remark, if $K \ge 0$, it is possible to
% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
% of the same program.  It can be done either by adapting the proof or by
% applying the {\sc L:Frame} rule one top of the triple already derived.
% More generally, without function calls, the {\sc L:Frame}
% rule is admissible in our system.

@@

Since we have a formal cost semantics, we can run our examples with
this semantics for different inputs and measure the cost to compare it
to our derived bound.  \pref{fig:3d} shows such a comparison for
example \emph{t08}.  One can see the derived constant factors are the
best possible if the input variable $x$ is non-negative.  If $x$ is
negative then the bound is only slightly off.

>>>>

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for Clight programs.
% The purpose of this section is to build some intuition that should
% make the following sections easy to read.

@@

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
quantity we are interested in.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that $(S, \state) \Downarrow_n \state'$.

To obtain a compositional reasnoning we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for the resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch}.% that holds for all
%stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
It holds for all stack metrics $M$.
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  That
is, we use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatedly increments a binary counter
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks
executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.  A
naive analysis of the algorithm would lead to a quadratic bound since
there can be a linear number of updates in the inner loop.

@@

See \pref{sec:logic} for more a formal definition of the logic and
additional example derivations.

@@

To enable automation, we fix the shape of the potential functions
$\Phi$ so that it becomes possible to use \emph{linear programming}
(LP) to compute a derivation in the logic.  If we assume for now that
a program state $\state$ simply maps variables to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$. Note
that there are no restrictions on the signs of the input variables in
the examples.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at the assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
that consumes $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $x + y + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variable for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear inequalities on
these variables that enforce that the potential annotations are sound.
For instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
of track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

@@

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource cost is closely related to the
number of times the test \code{a[i] > m} is true during the
execution.
%
If we define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
the number of ``maximum candidates'' in the array \code{a} seen
by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
So any automated tool would at best derive the linear bound $2 \cdot
N$ for that program.  But with the expressivity of our logic it is
possible to use the previous set cardinal directly and precisely tie
the bound to the initial contents of the array.
%
The non-trivial part of this derivation is finding the loop invariant
$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
When the condition \code{a[i] > m} is true, we know that we
encountered a ``maximum so far'' because $m$ is a maximal element of
\code{a[0 \dots i]}, thus $A(i) = 1 + A(i+1)$ and we get one
potential unit to pay for the assignment.  In the other case, no
maximum so far is encountered so $A(i) = A(i+1)$.
% As a side remark, if $K \ge 0$, it is possible to
% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
% of the same program.  It can be done either by adapting the proof or by
% applying the {\sc L:Frame} rule one top of the triple already derived.
% More generally, without function calls, the {\sc L:Frame}
% rule is admissible in our system.

@@

Since we have a formal cost semantics, we can run our examples with
this semantics for different inputs and measure the cost to compare it
to our derived bound.  \pref{fig:3d} shows such a comparison for
example \emph{t08}.  One can see that the derived constant factors are the
best possible if the input variable $x$ is non-negative.  If $x$ is
negative then the bound is only slightly off.
